---
title: "R Notebook"
author: "Owen O'Connor"
date: "May 10 2021"
output: word_document
---

## KAQI - Outdoor eggs
### Progress So Far

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(feather)
library(wesanderson)
library(PerformanceAnalytics)
library(openair)

###Prep Julie Data
mydir1 = "Julie_Outdoor_Data"
myfiles1 = list.files(path=mydir1, pattern="*.csv", full.names=TRUE)

julie_data <-myfiles1 %>%
  map_dfr(read_csv, col_types = "Tcdddddddddddddddddddd")    # read in all the files individually, using


#combine data from different csv files with differing column names
julie_data <- julie_data %>%
  unite("temp",`temperature[n/a]`,`temperature[degC]`, na.rm = TRUE) %>%
  unite("RH",`humidity[n/a]`,`humidity[percent]`, na.rm = TRUE) %>%
  unite("pressure",`pressure[n/a]`,`pressure[Pa]`, na.rm = TRUE) %>%
  unite("co2",`co2[n/a]`,`co2[ppm]`, na.rm = TRUE) %>%
  unite("pm10p0",`pm10p0[n/a]`,`pm10p0[ug/m^3]`, na.rm = TRUE) %>%
  unite("pm1p0",`pm1p0[ug/m^3]`,`pm1p0[ug/m^3]`, na.rm = TRUE) %>%
  unite("pm2p5",`pm2p5[n/a]`,`pm2p5[ug/m^3]`, na.rm = TRUE) 

julie_data <- julie_data %>%
  select(
    time, temp, RH, pressure, co2, pm10p0, pm1p0, pm2p5
  )
  
#convert columns to numeric
julie_data$temp <- as.numeric(julie_data$temp)
julie_data$RH <- as.numeric(julie_data$RH)
julie_data$pressure <- as.numeric(julie_data$pressure)
julie_data$co2 <- as.numeric(julie_data$co2)
julie_data$pm10p0 <- as.numeric(julie_data$pm10p0)
julie_data$pm1p0 <- as.numeric(julie_data$pm1p0)
julie_data$pm2p5 <- as.numeric(julie_data$pm2p5)
```

```{r}
### Prep Lorraine Data

mydir2 = "Lorraine_Outdoor_Data"
myfiles2 = list.files(path=mydir2, pattern="*.csv", full.names=TRUE)

lorraine_data <- myfiles2 %>%
  map_dfr(read_csv, col_types = "Tcdddddddddddddddddddd")    # read in all the files individually, using

#combine data from different csv files with differing column names
lorraine_data <- lorraine_data %>%
  unite("temp",`temperature[n/a]`,`temperature[degC]`, na.rm = TRUE) %>%
  unite("RH",`humidity[n/a]`,`humidity[percent]`, na.rm = TRUE) %>%
  unite("pressure",`pressure[n/a]`,`pressure[Pa]`, na.rm = TRUE) %>%
  unite("co2",`co2[n/a]`,`co2[ppm]`, na.rm = TRUE) %>%
  unite("pm10p0",`pm10p0[n/a]`,`pm10p0[ug/m^3]`, na.rm = TRUE) %>%
  unite("pm1p0",`pm1p0[ug/m^3]`,`pm1p0[ug/m^3]`, na.rm = TRUE) %>%
  unite("pm2p5",`pm2p5[n/a]`,`pm2p5[ug/m^3]`, na.rm = TRUE) 

lorraine_data <- lorraine_data %>%
  select(
    time, temp, RH, pressure, co2, pm10p0, pm1p0, pm2p5
  )

#convert columns to numeric
lorraine_data$temp <- as.numeric(lorraine_data$temp)
lorraine_data$RH <- as.numeric(lorraine_data$RH)
lorraine_data$pressure <- as.numeric(lorraine_data$pressure)
lorraine_data$co2 <- as.numeric(lorraine_data$co2)
lorraine_data$pm10p0 <- as.numeric(lorraine_data$pm10p0)
lorraine_data$pm1p0 <- as.numeric(lorraine_data$pm1p0)
lorraine_data$pm2p5 <- as.numeric(lorraine_data$pm2p5)

#read in the tibble prepared by andy.R
eggmasterk <- read_feather("eggmasterk.feather")

eggmasterk <- eggmasterk %>%
  select(
    `time`,`temperature[degC]`,`humidity[percent]`, `co2[ppm]`,`pm10p0[ug/m^3]`,`pm2p5[ug/m^3]`
  ) %>%
  rename(
    pm10p0 = `pm10p0[ug/m^3]`,
    pm2p5 = `pm2p5[ug/m^3]`,
    co2 = `co2[ppm]`,
    temp= `temperature[degC]`,
    RH = `humidity[percent]`
  ) 

eggmasterk$pm10p0 <- as.numeric(eggmasterk$pm10p0)
eggmasterk$pm2p5 <- as.numeric(eggmasterk$pm2p5)
eggmasterk$co2 <- as.numeric(eggmasterk$co2)
eggmasterk$temp <- as.numeric(eggmasterk$temp)
eggmasterk$RH <- as.numeric(eggmasterk$RH)

eggmasterk<-mutate(eggmasterk, time= parse_datetime(time))

```
Now, we will take a look at what we've succesfully loaded in. 


```{r}
julie_data %>% 
  rename(
    date = time
  ) %>%
  summaryPlot(period = "months")
```
```{r}

lorraine_data %>% 
  rename(
    date = time
  ) %>%
  summaryPlot(period = "months")
```
```{r}

eggmasterk %>% 
  na.omit() %>%
  rename(
    date = time
  ) %>%
  summaryPlot(period = "months")



```
Ok great. We've got some chunks missing here or there, and the eggs came on different days, but overall we've got the three eggs running outdoors at three different locations over roughly the same period. I want to better understand why the Andy Murphy egg has some many more observtions. Presummably it's set to take measurements more often, but let's check.

```{r}
eggmasterk[600:610,]
julie_data[600:610,]
lorraine_data[600:610,]


```
This confirms our assumption. The AndyMurphy egg is taking readings every minute, while the eggs at Julie and Lorraine are taking readings every 30 minutes. 
Now let's join the datasets together. 
First, let's group the Andy Murphy data by averaging the data into 30 minute chuncks so it lines up with our other eggs.
```{r}
eggmasterk <- eggmasterk[-1,] %>%
group_by(time = cut(time, breaks="30 min")) %>%
  summarize(
    temp = mean(temp, na.rm = TRUE),
    RH = mean(RH,na.rm = TRUE),
    co2 = mean(co2,na.rm = TRUE),
    pm10p0 = mean(pm10p0,na.rm = TRUE),
    pm2p5 = mean(pm2p5,na.rm = TRUE)
            ) 


eggmasterk <- mutate(eggmasterk, time = parse_datetime(as.character(time)))
  
eggmasterk[600:610,]

```

Join the three data sets. For now, we're taking only the observations where all three eggs are recording readings. 
```{r}
glimpse(eggmasterk)

outdoor <- inner_join(lorraine_data,julie_data,by="time", suffix = c(".lorraine", ".julie")) %>% 
  inner_join(eggmasterk,by="time") 
  #%>%
  #na.omit()

```

```{r}
glimpse(outdoor)
```

```{r}
outdoor %>% 
  rename(date = time) %>%
summaryPlot(period = "months")

```
Cool. So we've got the data of interest from the three eggs, and selected the observations where we have readings from all three eggs. That means we've lost the chunk in second half of June when the Andy Murphy egg was down. But overall, seems like we're doing pretty good. 

Let's plot the PM2.5 data from the three sensors over time:

```{r}
#Time Series plot
ggplot(outdoor, aes(time,y)) +
  geom_point(aes(y = pm2p5.lorraine), size= .3, color = "red", alpha =.4) +
  geom_point(aes(y = pm2p5.julie), size= .3, color = "blue", alpha =.4) +
  geom_point(aes(y = pm2p5), size= .3, color = "green", alpha =.4) +
  geom_smooth(aes(y = pm2p5.lorraine), color = "red", se =FALSE) +
  geom_smooth(aes(y = pm2p5.julie),color = "blue", se =FALSE) +
  geom_smooth(aes(y = pm2p5),color = "green", se =FALSE) +
  theme_classic() +
  labs(
    y = "pm2.5",
    x = NULL,
    title = paste(
      "Comparing Outdoor Eggs"
    ),
    subtitle = paste(
      "Andy = Green, Julie = Blue, Lorraine = Red"
    )) #+
#scale_y_continuous(limits = c(0,50))
```
The chart is hard to see because one day in early July is throwing off the scale with high readings from Lorraine's house and Andy Muprhy. Pretty sure that's July 4th, but haven't zoomed in yet. Anyway, let's limit the scale and see if we can see a little better:

```{r}
#Time Series plot
ggplot(outdoor, aes(time,y)) +
  geom_point(aes(y = pm2p5.lorraine), size= .3, color = "red", alpha =.4) +
  geom_point(aes(y = pm2p5.julie), size= .3, color = "blue", alpha =.4) +
  geom_point(aes(y = pm2p5), size= .3, color = "green", alpha =.4) +
  geom_smooth(aes(y = pm2p5.lorraine), color = "red", se =FALSE) +
  geom_smooth(aes(y = pm2p5.julie),color = "blue", se =FALSE) +
  geom_smooth(aes(y = pm2p5),color = "green", se =FALSE) +
  theme_classic() +
  labs(
    y = "pm2.5",
    x = NULL,
    title = paste(
      "Comparing Outdoor Eggs"
    ),
    subtitle = paste(
      "Andy = Green, Julie = Blue, Lorraine = Red"
    )) +
scale_y_continuous(limits = c(0,60))
```
I'd also like to zoom in on a week and see how these eggs track together. Let's do that:

```{r}
#Time Series plot
ggplot(subset(outdoor, month(time) == 11 & day(time) > 1 & day(time) < 10), aes(time,y)) +
  geom_point(aes(y = pm2p5.lorraine), size= .3, color = "red", alpha =.4) +
  geom_point(aes(y = pm2p5.julie), size= .3, color = "blue", alpha =.4) +
  geom_point(aes(y = pm2p5), size= .3, color = "green", alpha =.4) +
  #geom_smooth(aes(y = pm2p5.lorraine), color = "red", se =FALSE) +
  #geom_smooth(aes(y = pm2p5.julie),color = "blue", se =FALSE) +
  #geom_smooth(aes(y = pm2p5),color = "green", se =FALSE) +
  theme_classic() +
  labs(
    y = "pm2.5",
    x = NULL,
    title = paste(
      "Comparing Outdoor Eggs"
    ),
    subtitle = paste(
      "Andy = Green, Julie = Blue, Lorraine = Red"
    )) +
scale_y_continuous(limits = c(0,50))
```
Ok. Looks like they are pretty correlated. Speaking of correlation, let's do a correlation matrix:

```{r}
outdoor %>%
  select(
    pm2p5.lorraine,pm2p5.julie,pm2p5,pm10p0.lorraine,pm10p0.julie, pm10p0) %>%
  rename(
    pm2p5.andy = pm2p5,
    pm10p0.andy = pm10p0
    ) %>%
chart.Correlation()
```
Now, let's take a look at boxplots to see the distribution a little. We will look at them without outliers plotted, because the outliers throw off the scale:


```{r}
#preparing data for boxplots
boutdoor <- gather(outdoor,`Sensor`, pm2.5,c(pm2p5.julie,pm2p5.lorraine,pm2p5))%>%
  mutate(
    month = month(time, label=TRUE)
  )

ggplot(boutdoor,aes(x=`Sensor`,pm2.5, fill = `Sensor`)) +
  geom_boxplot(outlier.shape = NA) +
  stat_boxplot(geom = "errorbar") +
  stat_summary(fun=mean, geom="point", shape=20, size=3, color="yellow") +
  labs(
    y = expression(PM2.5 - (μg/~m^3)),
    x = NULL,
    title = paste(
      "2020 Particulate Matter - Kingston"
    ),
    subtitle = paste(
      "without outliers"
    )) +
  scale_fill_manual(labels = c("Andy","Julie","Lorraine"),values=wes_palette(n=3, name="Darjeeling1")) +
  scale_y_continuous(limits = c(0,30)) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5))
```
And broken down by month:

```{r}
ggplot(boutdoor,aes(x=month,pm2.5, fill = `Sensor`)) +
  geom_boxplot(outlier.shape = NA) +
  stat_boxplot(geom = "errorbar") +
  labs(
    y = expression(PM2.5 - (μg/~m^3)),
    x = NULL,
    title = paste(
      "2020 Particulate Matter - Kingston"
    ),
    subtitle = paste(
      "without outliers"
    )) +
  scale_fill_manual(labels = c("Andy","Julie","Lorraine"),values=wes_palette(n=3, name="Darjeeling1")) +
  scale_y_continuous(limits = c(0,60)) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5))

```
Looks like the eggs track pretty closely to each other. But Julie's egg seems like it is consistent reading higher PM2.5 than the other eggs. Likes drill into that a bit. 

First thought it to run an ANOVA test, but from the summaryPlot() above, it seems unlikely that pm2.5 is normally distributed. So we will use the non-parametric krushal-wallice test. Any better ideas? Would you rather the data be transformed and use ANOVA?


```{r}
library(rstatix)
egg_kruskal <- kruskal_test(pm2.5 ~`Sensor`,data=boutdoor)
egg_kruskal
```

Low p value shows that it's likely there is a statistically significant difference among the means three outdoor eggs (for PM2.5). Now lets see which pairs have a statistically significant differences. 

```{r}
wilcox <- pairwise.wilcox.test(boutdoor$pm2.5, boutdoor$Sensor,
                 p.adjust.method = "bonferroni")
wilcox
```

Looks like all of them show up as a statistically significant difference from each other. Judging by the boxplots, my first impression is that this is a statistically significant difference but not necessarily a biologically significant difference. 


What other questions are on your mind about comparing these three eggs? 

```

