---
title: "QuantAQ vs Profiler"
author: "Owen O'Connor"
date: "May 10 2021"
output: word_document
---

There was a short overlap in February 2021 where both the QuantAQ and the MetOne Profiler were running on top of Andy Murphy Center.
Objective is to compare the readings of these two Particulate Matter sensors for this period.
Loading in the data from the two profilers:
```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(feather)
library(wesanderson)
library(PerformanceAnalytics)

quantaq <- read_csv("8237c1fcb6fa40109fe06d06458df1ac.csv") %>%
  rename(
    Time = timestamp_local
  )


profiler <- read_csv("Copy of 01 2021-01-28-16-07-41.csv", skip = 2)

# Change time strings to datetime type
profiler <- mutate(profiler, Time= parse_date_time(Time,"mdYHM"))

profiler <- profiler %>%
  rename(
    c0.3 = `0.3 (Counts)`,
    c0.5 = `0.5 (Counts)`,
    c0.7 = `0.7 (Counts)`,
    c1.0 = `1.0 (Counts)`,
    c2.0 = `2.0 (Counts)`,
    c3.0 = `3.0 (Counts)`,
    c5.0 = `5.0 (Counts)`,
    c10.0 = `10.0 (Counts)`
  )

# Filtering for negative values for counts, and values over 900,000,000
profiler <- profiler %>% 
  filter(
    c0.3 >= 0 & c0.3 <= 900000000,
    c0.5 >= 0 & c0.5 <= 900000000,
    c0.7 >= 0 & c0.7 <= 900000000,
    c1.0 >= 0 & c1.0 <= 900000000,
    c2.0 >= 0 & c2.0 <= 900000000,
    c3.0 >= 0 & c3.0 <= 900000000,
    c5.0 >= 0 & c5.0 <= 900000000,
    c10.0 >= 0 & c10.0 <= 900000000,
    Alarms == 0
  )

# Using average diameters for each bin to get count totals by size
profiler <- profiler %>%
  mutate(
    tot0.4count = c0.3 - c0.5, 
    tot0.6count = c0.5 - c0.7,
    tot0.85count = c0.7 -c1.0, 
    tot1.5count = c1.0 - c2.0, 
    tot2.5count = c2.0 - c3.0, 
    tot4.0count = c3.0 - c5.0, 
    tot7.5count = c5.0 - c10.0,
    tot20count = c10.0
  )

# Sphere volume constants. For each bin, we're calculating the amount of 
#volume in one particle of the midpoint size of the bin. V = 4/3*pi*r^3
vol0.4 = (4/3) * pi * (.4/2)^3
vol0.6 = (4/3) * pi * (.6/2)^3
vol0.85 = (4/3) * pi * (.85/2)^3
vol1.5 = (4/3) * pi * (1.5/2)^3
vol2.5 = (4/3) * pi * (2.5/2)^3
vol4.0 = (4/3) * pi * (4.0/2)^3
vol7.5 = (4/3) * pi * (7.5/2)^3
vol20 = (4/3) * pi * (20/2)^3

# Calculating volumes
profiler <- profiler %>%
  mutate(
    totvol0.4 = (vol0.4*tot0.4count),
    totvol0.6 = (vol0.6*tot0.6count),
    totvol0.85 = (vol0.85*tot0.85count),
    totvol1.5 = (vol1.5*tot1.5count),
    totvol2.5 = (vol2.5*tot2.5count),
    totvol4.0 = (vol4.0*tot4.0count),
    totvol7.5 = (vol7.5*tot7.5count),
    totvol20 = (vol20*tot20count)
  )

# Estimating EPA thresholds
# 1)Sum the volume of particles of different sizes (in cubic microns) 
# 2)convert to cubic cm, dividing by 1000000000000
# 3) multiply by 1.65 to get the weight per cubic cm of particle in grams. 
# 4) Then multiply by 1000000 to convert mass in grams to micrograms
# 5) Convert the 1 lpm sampled to cubic meters (multiplying by 1000),
profiler <- profiler %>% rowwise() %>%
  mutate(
    pm2.5 = sum(c(totvol0.4,totvol0.6,totvol0.85,totvol1.5,totvol2.5)) / 1000000000000 * 1.65 *1000000 *1000,
    pm10 = sum(c(totvol0.4,totvol0.6,totvol0.85,totvol1.5,totvol2.5,totvol4.0,totvol7.5)) / 1000000000000 * 1.65 *1000000 *1000
  )

# Calculating mass for individual bins too, with the same formula that we used for PM2.5 and PM10
profiler <- profiler %>%
  mutate(
    mass0.4 =  totvol0.4 / 1000000000000 * 1.65 *1000000 *1000,
    mass0.6 =  totvol0.6 / 1000000000000 * 1.65 *1000000 *1000,
    mass0.85 =  totvol0.85 / 1000000000000 * 1.65 *1000000 *1000,
    mass1.5 =  totvol1.5 / 1000000000000 * 1.65 *1000000 *1000,
    mass2.5 =  totvol2.5 / 1000000000000 * 1.65 *1000000 *1000,
    mass4.0 =  totvol4.0 / 1000000000000 * 1.65 *1000000 *1000,
    mass7.5 =  totvol7.5 / 1000000000000 * 1.65 *1000000 *1000,
    mass20 =  totvol20 / 1000000000000 * 1.65 *1000000 *1000
  )


# Adjusting for Flow(lpm)
profiler <- profiler %>% 
  mutate(
    pm2.5 = pm2.5 * `Flow(lpm)`,
    pm10 = pm10  * `Flow(lpm)`
  )
```

Now we've got to join the sets together. I am removing all rows excpet where we have readings from both sensors in the same minute. (They are both taking a reading every minute)

```{r}

### Joining --------------------------------------

#reformatting quantAQ times - probably an easier way to do this
quantaq$datetime <- date(quantaq$Time)
hour(quantaq$datetime) <- hour(quantaq$Time)
minute(quantaq$datetime) <- minute(quantaq$Time)
quantaq$Time <- quantaq$datetime




full <- full_join(quantaq,profiler, by = "Time") 




cleanfull <- full[order(full$Time),] %>%
  select(
    Time, pm25, pm2.5, pm10.x, pm10.y
  ) %>%
  rename(
    pm2.5_quantaq = pm25,
    pm2.5_profiler = pm2.5,
    pm10_quantaq = pm10.x,
    pm10_profiler = pm10.y
  ) %>%
  na.omit()

```

Now we can see both Sensors plotted together. First PM2.5:

```{r}

#both together
ggplot(subset(cleanfull, month(Time) == 2 & day(Time) > 10 & day(Time) < 20),
       aes(x =Time, y = y)) +
  geom_point(aes(y = pm2.5_quantaq), color = "red", size= .3) +  #quantaq
  geom_point(aes(y = pm2.5_profiler), color = "blue", size = .3) + #profiler
  #scale_y_continuous(limits = c(0,25)) +
  labs(
    y = expression(Mass - (μg/~m^3)),
    x = NULL,
    title = paste(
      "Particulate Matter - Kingston"
    ),
    subtitle = paste(
      "PM2.5: Red = QuantAQ, Blue = Profiler "
    )) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5))
```
The outliers throw off the scale, so let's zoom in to see the variation in PM2.5 better:

```{r}

#both together
ggplot(subset(cleanfull, month(Time) == 2 & day(Time) > 10 & day(Time) < 20),
       aes(x =Time, y = y)) +
  geom_point(aes(y = pm2.5_quantaq), color = "red", size= .3) +  #quantaq
  geom_point(aes(y = pm2.5_profiler), color = "blue", size = .3) + #profiler
  scale_y_continuous(limits = c(0,25)) +
  labs(
    y = expression(Mass - (μg/~m^3)),
    x = NULL,
    title = paste(
      "Particulate Matter - Kingston"
    ),
    subtitle = paste(
      "PM2.5: Red = QuantAQ, Blue = Profiler "
    )) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5))
```




Now both sensors with PM10:

```{r}
#pm10 - both
ggplot(subset(cleanfull, month(Time) == 2 & day(Time) > 10 & day(Time) < 20),
       aes(x =Time, y = y)) +
  geom_point(aes(y = pm10_quantaq), color = "red", size= .3) +  #quantaq
  geom_point(aes(y = pm10_profiler), color = "blue", size = .3) + #profiler
  #scale_y_continuous(limits = c(0,50)) +
  labs(
    y = expression(Mass - (μg/~m^3)),
    x = NULL,
    title = paste(
      "Particulate Matter - Kingston"
    ),
    subtitle = paste(
      "PM10: Red = QuantAQ, Blue = Profiler "
    )) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5))

```

But lets zoom in on the Y axis to see variation in PM10 better;

```{r}
#pm10 - both
ggplot(subset(cleanfull, month(Time) == 2 & day(Time) > 10 & day(Time) < 20),
       aes(x =Time, y = y)) +
  geom_point(aes(y = pm10_quantaq), color = "red", size= .3) +  #quantaq
  geom_point(aes(y = pm10_profiler), color = "blue", size = .3) + #profiler
  scale_y_continuous(limits = c(0,50)) +
  labs(
    y = expression(Mass - (μg/~m^3)),
    x = NULL,
    title = paste(
      "Particulate Matter - Kingston"
    ),
    subtitle = paste(
      "PM10: Red = QuantAQ, Blue = Profiler "
    )) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5))


```

To check correlation, we will select certain variables and make a correlation matrix:

```{r}

library(PerformanceAnalytics)
cleanfull %>%
  select(
    -Time) %>%
  chart.Correlation()
```



